Multiway Spherical Clustering via \\ Degree-Corrected Tensor Block Models

### 2022-01-02 tensor_clustering/tensor_clustering_ieee/ieee.tex
update

\documentclass[lettersize,journal]{IEEEtran}
\usepackage{amsmath,amsfonts}
%\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{array}
\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\usepackage{textcomp}
\usepackage{stfloats}
\usepackage{url}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{cite}
\hyphenation{op-tical net-works semi-conduc-tor IEEE-Xplore}
% updated with editorial comments 8/9/2021

@@

\usepackage{amsmath,amssymb,amsthm,bm,mathtools}
\usepackage{algorithm}
\usepackage{dsfont,multirow,hyperref,setspace,enumerate}
\hypersetup{colorlinks,linkcolor=blue,citecolor=blue,urlcolor={black}}
\usepackage{lscape}
\usepackage{afterpage}
\usepackage{hyperref}

@@

@@

\textbf{Other conference paper to journal}
\begin{enumerate}
    \item Garvesh. Add more discussion about Theorem 5 and empirical simulations.
    %\item Eric Chi. Add one more remakr; One more section about the numerical issue; More comparison with other methods; add more future discussion.
    \item Anru. Add application in introduction; more remarks for algorithm and theorems; One more section about the application to a specific model; more experiments and discussions.
\end{enumerate}

@@

\fixme{Jiaxin}{Change the wording: \emph{individual/block effects} $\rightarrow$ \emph{individual/block parameters}, \emph{degree heretogeneity} $\rightarrow$ \emph{non-equal degrees}.}

@@

Classical tensor block model suffers from drawbacks to model real world data in spite of the popularity. The key underlying assumption of block model is that all nodes in the same community are exchangeable; i.e., the nodes have no individual effects apart from the block effects. However, the exchangeability assumption is often non-realistic. Each node may contribute to the data variation by its own multiplicative effect. Such degree heterogeneity appears commonly in social networks. For example, regular block model fails to model the Karate Club network~\citep{bickel2009nonparametric} without addressing
%accounting for 
degree heterogeneity. 
\fixme{Jiaxin}{Non-degree model ignoring the individual heterogeneity may seriously mislead the clustering results. Karate Club data analysis shows that clustering without heterogeneity tends to assign membership based on the degree of communication (two clusters for leaders and non-leaders) rather than the affiliation (two clusters for people from different clubs). }

The \emph{degree-corrected tensor block model} (dTBM) has been proposed recently to account for the degree heterogeneity~\citep{ke2019community}. The dTBM combines a higher-order checkerboard structure with degree parameter $\mtheta=(\mtheta(1),\ldots,\mtheta(p))^T$ to allow heterogeneity among $p$ nodes.  Figure~\ref{fig:intro} compares the underlying structures of TBM and dTBM with the same number of communities. The dTBM allows varying values within the same community, thereby allowing a richer structure.

{
\color{red}
We use the idea of \emph{spherical clustering} in initialization and iteration. Using angle-based similarity avoids the estimation of $\mtheta$.  
}

@@

{\bf Related work.} %Our work is closely related to, but also distinct from several lines of existing research. 
We emphasize the comparisons that set our work apart from earlier literature from two perspectives---models and algorithms. From the model perspective, our work extends the previous degree-corrected model from matrices to tensors. There is a huge literature on degree-corrected matrix models; see the review paper~\citep{abbe2017community} and the references therein. The tensor counterparts, however, are relatively less understood. Table~\ref{tab:comp} summarizes  the most relevant models for higher-order clustering. Earlier tensor methods either fail to allow degree heterogeneity~\citep{chien2018community,han2020exact,zhang2021exact,ghoshdastidar2017uniform,kim2018stochastic}, or suffer from sub-optimal misclustering rates~\citep{ke2019community,ghoshdastidar2017consistency,chi2020provable}. In contrast, our method addresses the degree heterogeneity, allows discrete and continuous entries, and achieves exponentially fast rate in clustering tasks.

From the algorithm perspective, our methods generalize the recent global-to-local strategy for matrix learning~\citep{gao2018community,chi2019nonconvex,yun2016optimal} to tensors~\citep{han2020exact,ahn2018hypergraph,kim2018stochastic}. Despite the conceptual similarity, we address several fundamental challenges associated with this non-convex, non-continuous problem. We show the insufficiency of the conventional tensor HOSVD~\citep{kolda2009tensor}, and we develop a weighted higher-order initialization that relaxes the eigen-gap separation condition. Furthermore, our local iteration leverages the angle-based clustering in order to avoid explicit estimation of degree heterogeneity. Our bounds reveal the interesting interplay between the computational and statistical errors. We show that our final estimate provably achieves the exact clustering within only polynomial-time complexity.

Our work is closely related to, but also distinct from several lines of existing research:
\begin{enumerate}

\item \textbf{Degree-corrected model.}
    \begin{itemize}
        \item Matrix. (1) Previous work, survey paper (2) Unique technical challenges in Tensor (3) matrix does not show statistical-computational gap;
        \item Tensor. (Comparison with Ke) (1) Ke does not provide the SNR limits (2) Ke's algorithm is sub-optimal (3) Ke focuses on Binary data. 
    \end{itemize}

    \item \textbf{Higher-order clustering.} (1) Previous work fails to account the heterogeneity, which is not realistic; (2) The extension of theoretical results highly challenging.  Extra nuisance parameter makes it more difficult to establish the statistical and computational limits.

    \item \textbf{Spherical clustering.}

    \item \textbf{Global-to-local algorithm.} (1) Matrix global-to-local algorithm (2) Tensor global-to-local algorithm is used for non-degree algorithm. Extra nuisance parameter makes it necessary to consider angle-based similarity; angle-based clustering makes the Cartesian coordinates based analysis non-applicable for our setting. 
\end{enumerate}

\begin{table*}[t]
\resizebox{\textwidth}{!}{%
    \begin{tabular}{c|ccccc}
\hline
    & \cite{gao2018community}&  \cite{han2020exact}& \cite{ke2019community}&  \cite{ghoshdastidar2017consistency} & \textbf{Ours}\\
    \hline
        {\small Allow tensors}& $\times$ & $\surd$ & $\surd$ &$\surd$ & $\surd$  \\
        {\small Allow heterogeneity} &  $\surd$ & $\times$ & $\surd$ &$\surd$ & $\surd$ \\
        %Assortative assumption&  $\surd$ & $\times$ & $\times$ & $\times$\\
       {\small Allow various data} & $\times$ & $\surd$ & $\times$ & $\surd$  &$\surd$\\
        {\small Misclustering rate (for order 2$^*$)}& $\exp(-p)$ & $\exp(-p)$ & $p^{-1}$ & $p^{-1}$& $\exp(-p)$\\
        \hline
    \end{tabular}
    }
    \caption{Comparison between previous methods with our method. $^*$We list here only the result for order-2 tensors (i.e., matrix). Our results allow general tensors of arbitrary order $K$; See Theorem~\ref{thm:refinement}.}\label{tab:comp}
\end{table*}

{
\color{red}
Organization.
}

@@

@@

Suppose we have an order-$K$ data tensor $\tY \in \bbR^{p \times \cdots \times p}$. For ease of notation, we focus on symmetric tensors in this section; our framework easily extends to general asymmetric tensors. Assume there exist $r \geq 2$  disjoint communities among the $p$ nodes. We represent the community assignment by a function $z \colon [p]\mapsto[r]$, where $z(i) = a$ for $i$-th node that belongs to the $a$-th community. Then, $z^{-1}(a)=\{i\in[p]\colon z(i)=a\}$ denotes the set of nodes that belong to the $a$-th community, and $|z^{-1}(a)|$ denotes the number of nodes in the $a$-th community.
%, and we let $p_a = \sum_{i \in [p]}\ind \offf{ z_i= a}$ denote the size of $a$-th community. 
Let $\mtheta=(\theta(1),\ldots,\theta(p))^T$ denote the degree heterogeneity for $p$ nodes. We consider the order-$K$ dTBM~\citep{ghoshdastidar2017consistency,ke2019community},
\begin{equation}\label{eq:model_margin}
    \tY(i_1,\ldots,i_K) = \tS(z(i_1),\ldots,z(i_K)) \prod_{k = 1}^K\theta_{i_k} + \tE(i_1,\ldots,i_K), 
\end{equation}
\normalsize
where $\tS \in \bbR^{r \times \cdots \times r}$ is an order-$K$ tensor collecting the block means among communities, and $\tE\in\mathbb{R}^{p\times \cdots \times p}$ is a noise tensor consisting of independent mean-zero sub-Gaussian entries with variance bounded by $\sigma^2$. The unknown parameters are $z$, $S$, and $\mtheta$. The dTBM can be equivalently written in a compact form of tensor-matrix product:
\begin{equation}\label{eq:model_tensor}
 \mathbb{E}\tY = \tS \times_1 \mTheta \mM \times_2 \cdots \times_K  \mTheta \mM,
\end{equation}
where $\mTheta = \text{diag}(\theta(1),...,\theta(p)) \in \bbR^{p \times p}$ is a diagonal matrix, $\mM \in \offf{0,1}^{p \times r}$ is the membership matrix associated with community assignment $z$ such that $\mM(i,j)=\ind\{z(i)=j\}$. By definition, each row of $\mM$ has one copy of 1's and 0's elsewhere. Note that the discrete nature of $\mM$ renders our model~\eqref{eq:model_tensor} more challenging than Tucker decomposition. %The Tucker rank of the signal tensor $\bbE \tY$ is bounded by, but not necessarily equal to, $(r,...,r)$~\citep{wang2019multiway}. 
We call a tensor $\tX$ an $r$-block tensor with degree $\mtheta$
%with the core tensor $\tS$ 
if $\tX$ admits dTBM~\eqref{eq:model_tensor}. Here, we give two special cases of dTBM.
\begin{example}[Gaussian TBM]
Let $\theta(i)= 1$ for all $i \in [p]$ and $\tE$ be a noise tensor with i.i.d.\ $N(0,\sigma^2)$ entries. Our dTBM reduces to a non-degree Gaussian TBM, which is widely used in previous clustering algorithms~\citep{wang2019multiway,chi2020provable,han2020exact}. The theoretical results in TBM serve as benchmarks for dTBM. 
\end{example}

\begin{example}[Binary dTBM]
Consider a $K$-uniform hypergraph $H = (V,E)$, where $V = [p]$ collects the nodes with $r$ disjoint communities and $E$ collects all the $K$-way hyperedges. Let $\tY \in \{0,1\}^{p \times \cdots \times p}$ denote the adjacency tensor, where the entries encode the presence or absence of hyperdeges among $p$ nodes. Specifically, let $\tY(i_1,...,i_K) = 1$ if  $(i_1,...,i_K) \in E$, otherwise, $\tY(i_1,...,i_K) = 0$, for all $(i_1,\ldots,i_K) \in [p]^K$. The equation~\eqref{eq:model_tensor} models $\mathbb{E}\tY$ with unknown degree heterogeneity and subgaussianity parameter $\sigma^2 = 1/4$. 
\end{example}

@@

\subsection{Identifiability under angle gap condition}
The goal of clustering is to estimate the partition function $z$ from model~\eqref{eq:model_tensor}. We use $\tP$ to denote the following parameter space for $(z,\tS,\mtheta)$,
\footnotesize
\begin{align}\label{eq:family}
\tP&=\bigg\{(z,\tS,\mtheta)\colon  \ \mtheta\in\mathbb{R}^p_{+},\ 
{c_1p\over r}\leq |z^{-1}(a)|\leq {c_2 p\over r},\\& c_3\leq \onormSize{}{\text{Mat}(\tS)_{a:}}\leq c_4, \onorm{\mtheta_{z^{-1}(a)}}_1=|z^{-1}(a)|, a\in[r]\bigg\},
\end{align}
\normalsize
where $c_i>0$'s are universal constants.  
%We briefly describe the rationale of the constraints in~\eqref{eq:family}. 
First, the entrywise positivity constraint on  $\mtheta\in\mathbb{R}^p_{+}$ is imposed to avoid sign ambiguity between entries in $\mtheta_{z^{-1}(a)}$ and $\tS$. This constraint can be achieved without sacrificing model flexibility, by using a slightly larger dimension of $\tS$ in the factorization~\eqref{eq:model_tensor}; see Example~\ref{ex:positive}. Second, 
%the quantity $|z^{-1}(a)|$ denotes the number of nodes in $a$-th community. T
the constants $c_1, c_2$ in the $|z^{-1}(a)|$ bound assume the roughly balanced size across $r$ communities. Third, the constants $c_3, c_4 $ in the magnitude of $\text{Mat}(\tS)_{a:}$ requires no purely zero slide in $\tS$, so the core tensor $\tS$ is not trivially reduced to lower rank. Lastly, the $\ell_1$ normalization $\onormSize{}{\mtheta_{z^{-1}(a)}}_1=|z^{-1}(a)|$ is imposed to avoid the scalar ambiguity between $\mtheta_{z^{-1}(a)}$ and $\tS$. This constraint, again, incurs no restriction to model flexibility but makes our presentation cleaner.

>>>>

\documentclass[lettersize,onecolumn,journal]{IEEEtran}
\usepackage{amsmath,amsfonts}
%\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{array}
\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\usepackage{textcomp}
\usepackage{stfloats}
\usepackage{url}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{cite}
\hyphenation{op-tical net-works semi-conduc-tor IEEE-Xplore}
% updated with editorial comments 8/9/2021

@@

\usepackage{amsmath,amssymb,amsthm,bm,mathtools}
\usepackage{algorithm}
\usepackage{dsfont,multirow,hyperref,setspace,enumerate}
\hypersetup{colorlinks,linkcolor=black,citecolor=black,urlcolor={black}}
\usepackage{lscape}
\usepackage{afterpage}
\usepackage{hyperref}

@@

\usepackage{tabularx,booktabs}
\newcolumntype{Y}{>{\centering\arraybackslash}X}

@@

% {
% \color{red}

% \textbf{Other conference paper to journal}
% \begin{enumerate}
%     \item Garvesh. Add more discussion about Theorem 5 and empirical simulations.
%     %\item Eric Chi. Add one more remakr; One more section about the numerical issue; More comparison with other methods; add more future discussion.
%     \item Anru. Add application in introduction; more remarks for algorithm and theorems; One more section about the application to a specific model; more experiments and discussions.

%     \item Christina. Almost re-write the paper. 
% \end{enumerate}
% }

% {
% \color{blue}

% \subsection{Our contribution.} 
% \begin{itemize}
%     \item We propose a generalized dTBM.
%     \item Phase transition.
%     \item We propose a multiway spherical clustering algorithm.
% \end{itemize}

% \subsection{Related work.} We emphasize the comparisons that set our work apart from earlier literature from two perspectives -- models and algorithms.

% From the model perspective, our proposed model is related, but also distinct to following lines of works.
% \begin{itemize}

%     %\item \textit{Tensor block model.}

%     \item \textit{Degree-corrected block model.}

%     \item \textit{Degree-corrected block model for hypergraphs.}

%     \item \textit{Other higher-order clustering models.}
% \end{itemize}

% From the algorithm perspective, our work is inspired by two key algorithmic techniques.

% \begin{itemize}
%     \item \textit{Spherical clustering.}

%     \item \textit{Global-to-local strategy.}
% \end{itemize}

% \subsection{Binary multiway spherical clustering.}

% \paragraph{Algorithm extension}

% \paragraph{Theoretical guarantee}

@@

%\fixme{Jiaxin}{Change the wording: \emph{individual/block effects} $\rightarrow$ \emph{individual/block parameters}, \emph{degree heretogeneity} $\rightarrow$ \emph{non-equal degrees}.}

@@

Classical tensor block model suffers from drawbacks to model real world data in spite of the popularity. The key underlying assumption of block model is that all nodes in the same community are exchangeable; i.e., the nodes have no individual effects apart from the block effects. However, the exchangeability assumption is often non-realistic. Each node may contribute to the data variation by its own multiplicative effect. Such degree heterogeneity appears commonly in social networks. {\color{blue} Ignoring the individual heterogeneity may seriously mislead the clustering results.} For example, regular block model fails to model {\color{blue} the member affiliation in} the Karate Club network~\citep{bickel2009nonparametric} without addressing
%accounting for 
degree heterogeneity. 
%\fixme{Jiaxin}{Non-degree model ignoring the individual heterogeneity may seriously mislead the clustering results. Karate Club data analysis shows that clustering without heterogeneity tends to assign membership based on the degree of communication (two clusters for leaders and non-leaders) rather than the affiliation (two clusters for people from different clubs). }

The \emph{degree-corrected tensor block model} (dTBM) has been proposed recently to account for the degree heterogeneity~\citep{ke2019community}. The dTBM combines a higher-order checkerboard structure with degree parameter $\mtheta=(\mtheta(1),\ldots,\mtheta(p))^T$ to allow heterogeneity among $p$ nodes.  Figure~\ref{fig:intro} compares the underlying structures of TBM and dTBM with the same number of communities. The dTBM allows varying values within the same community, thereby allowing a richer structure. {\color{blue} To solve dTBM, we project clustering objects to a unit sphere and consider the angle similarity; detailed algorithms are in Section~\ref{sec:alg}. On one hand, the \textit{spherical} clustering avoids the estimation of nuisance degree heterogeneity. On the other hand, the usage of angle similarity brings new challenges to the derivations of theoretical results, and the polar coordinates based techniques are equipped in the proof.
}

% {
% \color{red}
% We use the idea of \emph{spherical clustering} in initialization and iteration. Using angle-based similarity avoids the estimation of $\mtheta$.  
% }

@@

{\bf Related work.} 
{\color{blue}
Our work is closely related to, but also distinct from several lines of existing research. 
\begin{itemize}
    \item \textit{Block model for clustering.} Block models such as stochastic block model (SBM) and degree-corrected SBM has been widely used for matrix clustering problems. The theoretical properties and algorithm performance for block models have been well-studied; see the review paper~\citep{abbe2017community} and the references therein. However, The tensor counterparts are relatively less understood. Table~\ref{tab:comp} summarizes the most relevant models. Specifically,
    \begin{itemize}
        \item the tensor block model (TBM, \cite{wang2019multiway, han2020exact}), a higher-order extension of SBM, fails to allow degree heterogeneity. The Cartesian coordinates based analysis in \cite{han2020exact} is also non-applicable to handle the extra flexibility brought from heterogeneity. In contrast, our model addresses the degree heterogeneity, and the polar coordinates based tools are adapted for the theoretical analysis;
        \item the hypergraph degree-corrected block model (hDCBM) proposed by \cite{ke2019community, yuan2018testing} accounts for degree heterogeneity. Whereas, the hDCBM is designed only for binary observations, and the proposed spectral algorithm in \cite{ke2019community} achieves a sub-optimal clustering accuracy in higher-order scenarios. In constrast, our model allows discrete and continuous entries, and achieves exponentially fast rate in clustering tasks. More importantly, to our best knowledge, we are the first to provide the statistical and computational limits analyses for the degree-corrected block model in tensor clustering.
    \end{itemize}

    \item \textit{Global-to-local algorithm strategy.} Our methods generalize the recent global-to-local strategy for matrix learning~\citep{gao2018community,chi2019nonconvex,yun2016optimal} to tensors~\citep{han2020exact,ahn2018hypergraph,kim2018stochastic}. Despite the conceptual similarity, we address several fundamental challenges associated with this non-convex, non-continuous problem. We show the insufficiency of the conventional tensor HOSVD~\citep{kolda2009tensor}, and we develop a weighted higher-order initialization that relaxes the eigen-gap separation condition. Furthermore, our local iteration leverages the angle-based clustering in order to avoid explicit estimation of degree heterogeneity. Our bounds reveal the interesting interplay between the computational and statistical errors. We show that our final estimate provably achieves the exact clustering within only polynomial-time complexity.

\end{itemize}
}

%We emphasize the comparisons that set our work apart from earlier literature from two perspectives---models and algorithms. From the model perspective, our work extends the previous degree-corrected model from matrices to tensors. There is a huge literature on degree-corrected matrix models; see the review paper~\citep{abbe2017community} and the references therein. The tensor counterparts, however, are relatively less understood. Table~\ref{tab:comp} summarizes  the most relevant models for higher-order clustering. Earlier tensor methods either fail to allow degree heterogeneity~\citep{chien2018community,han2020exact,zhang2021exact,ghoshdastidar2017uniform,kim2018stochastic}, or suffer from sub-optimal misclustering rates~\citep{ke2019community,ghoshdastidar2017consistency,chi2020provable}. In contrast, our method addresses the degree heterogeneity, allows discrete and continuous entries, and achieves exponentially fast rate in clustering tasks.

%From the algorithm perspective, our methods generalize the recent global-to-local strategy for matrix learning~\citep{gao2018community,chi2019nonconvex,yun2016optimal} to tensors~\citep{han2020exact,ahn2018hypergraph,kim2018stochastic}. Despite the conceptual similarity, we address several fundamental challenges associated with this non-convex, non-continuous problem. We show the insufficiency of the conventional tensor HOSVD~\citep{kolda2009tensor}, and we develop a weighted higher-order initialization that relaxes the eigen-gap separation condition. Furthermore, our local iteration leverages the angle-based clustering in order to avoid explicit estimation of degree heterogeneity. Our bounds reveal the interesting interplay between the computational and statistical errors. We show that our final estimate provably achieves the exact clustering within only polynomial-time complexity.

% {
% \color{red}

% Our work is closely related to, but also distinct from several lines of existing research:
% \begin{enumerate}

% \item \textbf{Degree-corrected model.}
%     \begin{itemize}
%         \item Matrix. (1) Previous work, survey paper (2) Unique technical challenges in Tensor (3) matrix does not show statistical-computational gap;
%         \item Tensor. (Comparison with Ke) (1) Ke does not provide the SNR limits (2) Ke's algorithm is sub-optimal (3) Ke focuses on Binary data. 
%     \end{itemize}

%     \item \textbf{Higher-order clustering.} (1) Previous work fails to account the heterogeneity, which is not realistic; (2) The extension of theoretical results highly challenging.  Extra nuisance parameter makes it more difficult to establish the statistical and computational limits.

%     \item \textbf{Spherical clustering.}

%     \item \textbf{Global-to-local algorithm.} (1) Matrix global-to-local algorithm (2) Tensor global-to-local algorithm is used for non-degree algorithm. Extra nuisance parameter makes it necessary to consider angle-based similarity; angle-based clustering makes the Cartesian coordinates based analysis non-applicable for our setting. 
% \end{enumerate}

\begin{table*}[t]
\resizebox{\textwidth}{!}{%
\color{blue}
    \begin{tabular}{c|ccccc}
\hline
    & \cite{gao2018community}&  \cite{han2020exact}&  \cite{ghoshdastidar2017consistency} &\cite{ke2019community} & \textbf{Ours}\\
    \hline
        {\small Allow tensors}& $\times$ & $\surd$ & $\surd$ &$\surd$ & $\surd$  \\
        {\small Allow heterogeneity} &  $\surd$ & $\times$ & $\surd$ &$\surd$ & $\surd$ \\
        %Assortative assumption&  $\surd$ & $\times$ & $\times$ & $\times$\\
       %{\small Allow various data} & $\times$ & $\surd$ & $\times$ & $\surd$  &$\surd$\\
        {\small Eigen free clustering} &  $\surd$ & $\surd$ & $\times$ & $\times$ &$\surd$ \\
        %{\small Misclustering rate (for order 2$^*$)}& $\exp(-p)$ & $\exp(-p)$ & $p^{-1}$ & $p^{-1}$& $\exp(-p)$\\
        {\small Misclustering rate (for order $K^*$)}& - & $\exp(-p^{K/2})$ & $p^{-1}$& $p^{-2}$ &  $\exp(-p^{K/2})$\\
        \hline
    \end{tabular}
    }
    \caption{Comparison between previous methods with our method. $^*$We list the result for order-K tensors with $K \geq 3$ and general number of communities $r = \tO(1)$.
   % Our results allow general tensors of arbitrary order $K$; See Theorem~\ref{thm:refinement}.
    }\label{tab:comp}
\end{table*}

@@

{
\color{blue}

{\bf Organization.} The rest of this paper is organized as follows. Section~\ref{sec:model} introduces the degree-corrected tensor block model (dTBM) with three motivating examples and presents the identifiability of dTBM under the angle gap condition. We show the phase transition and the existence of statistical-computational gaps for the higher-order dTBM in Section~\ref{sec:limits}. In Section~\ref{sec:alg}, we provide a polynomial-time two-stage algorithm with misclustering rate guarantees. Numerical studies including the simulations to assess the theoretical results, comparison with other methods, and real data analysis on human brain connectome data and Peru legislation data are in Section~\ref{sec:simulation}. Last, we conclude our paper in Section~\ref{sec:con}.

@@

Suppose we have an order-$K$ data tensor $\tY \in \bbR^{p \times \cdots \times p}$. For ease of notation, we focus on symmetric tensors in this section; our framework easily extends to general asymmetric tensors. Assume there exist $r \geq 2$  disjoint communities among the $p$ nodes. We represent the community assignment by a function $z \colon [p]\mapsto[r]$, where $z(i) = a$ for $i$-th node that belongs to the $a$-th community. Then, $z^{-1}(a)=\{i\in[p]\colon z(i)=a\}$ denotes the set of nodes that belong to the $a$-th community, and $|z^{-1}(a)|$ denotes the number of nodes in the $a$-th community.
%, and we let $p_a = \sum_{i \in [p]}\ind \offf{ z_i= a}$ denote the size of $a$-th community. 
Let $\mtheta=(\theta(1),\ldots,\theta(p))^T$ denote the degree heterogeneity for $p$ nodes. We consider the order-$K$ dTBM~\citep{ghoshdastidar2017consistency,ke2019community},
\begin{equation}\label{eq:model_margin}
    \tY(i_1,\ldots,i_K) = \tS(z(i_1),\ldots,z(i_K)) \prod_{k = 1}^K\theta_{i_k} + \tE(i_1,\ldots,i_K), 
\end{equation}
\normalsize
where $\tS \in \bbR^{r \times \cdots \times r}$ is an order-$K$ tensor collecting the block means among communities, and $\tE\in\mathbb{R}^{p\times \cdots \times p}$ is a noise tensor consisting of independent mean-zero sub-Gaussian entries with variance bounded by $\sigma^2$. The unknown parameters are $z$, $S$, and $\mtheta$. The dTBM can be equivalently written in a compact form of tensor-matrix product:
\begin{equation}\label{eq:model_tensor}
 \mathbb{E}\tY = \tS \times_1 \mTheta \mM \times_2 \cdots \times_K  \mTheta \mM,
\end{equation}
where $\mTheta = \text{diag}(\theta(1),...,\theta(p)) \in \bbR^{p \times p}$ is a diagonal matrix, $\mM \in \offf{0,1}^{p \times r}$ is the membership matrix associated with community assignment $z$ such that $\mM(i,j)=\ind\{z(i)=j\}$. By definition, each row of $\mM$ has one copy of 1's and 0's elsewhere. Note that the discrete nature of $\mM$ renders our model~\eqref{eq:model_tensor} more challenging than Tucker decomposition. %The Tucker rank of the signal tensor $\bbE \tY$ is bounded by, but not necessarily equal to, $(r,...,r)$~\citep{wang2019multiway}. 
We call a tensor $\tX$ an $r$-block tensor with degree $\mtheta$
%with the core tensor $\tS$ 
if $\tX$ admits dTBM~\eqref{eq:model_tensor}. Here, we give two special cases of dTBM.

{
\color{blue}

\subsection{Motivating Examples.} Here, we provide three applications to illustrate the practical necessity of dTBM.

\vspace{0.2cm}
\paragraph{Tensor block model} Consider the model~\eqref{eq:model_tensor}. Let $\theta(i)=1$ for all $ i \in [p]$. The model~\eqref{eq:model_tensor} reduces to the tensor block model, which is widely used in previous clustering algorithms~\citep{wang2019multiway,chi2020provable,han2020exact}. The theoretical results in TBM serve as benchmarks for dTBM.

\vspace{0.2cm}

\paragraph{Community detection in hypergraphs} Hypergraph network is a powerful tool to collect the complex entity relations with higher-order interactions~\citep{ke2019community}. A typical undirected hypergraph is denoted as $H = (V,E)$, where $V = [p]$ is the set of nodes and $E$ is the set of undirected hyperedges. Each hyperedge in $E$ is a subset of $V$, and we call the hyperedge an order-$K$ edge if the corresponding subset involves $K$ nodes. We call $H$ a $K$-uniform hypergraph if $E$ only contains order-$K$ edges.

Similar with the adjacency matrix, it is natural to represent the $K$-uniform hypergraph by a binary order-$K$ adjacency tensor. Let $\tY \in \{0,1\}^{p \times \cdots \times p}$ denote the adjacency tensor, where the entries encode the presence or absence of hyperdeges among $p$ nodes. Specifically, 
\begin{equation}
    \tY(i_1,...,i_K) =  \begin{cases}
    1  & \text{if }  (i_1,...,i_K) \in E\\
    0 & \text{if }  (i_1,...,i_K) \notin E
    \end{cases},
\end{equation}
for all $(i_1,\ldots,i_K) \in [p]^K$.

Assume there exist $r$ disjoint communities among $p$ nodes. The equation~\eqref{eq:model_tensor} models $\mathbb{E}\tY$ with unknown degree heterogeneity $\mtheta$ and subgaussianity parameter $\sigma^2 = 1/4$.

\vspace{0.2cm}
\paragraph{Multi-layer weighted network} 
Multi-layer weighted network data consists of multiple networks with the same set of nodes. One representative example is the brain structural connectome data~\citep{zhang2019tensor}. The multi-layer weighted network $\tY$ has dimension of $p \times p \times L$, where $p$ denotes the number of brain regions in interest, and $L$ denotes the number of layer. Each of the $L$ networks describes one aspect of the brain connections, and the tensor $\tY$ include a mixture slices with continuous, binary, and count entries.

% We use $G = (V, \{E_l\}_{l=1}^L)$ denote a $L$-layer undirected multi-layer network, where $E_l$ is the set of edges in the network of the $l$-th layer. Let $\mY_l \in \{0,1\}^{p\times p}$ denote the adjacency matrix for the $L$-th layer with nodes $V$ and edges $E_l$. Then, the multi-layer network is represented by an order-3 binary tensor $\tY \in \{0,1\}^{p \times p \times L}$, where 
% \begin{equation}
%     \tY(i_1,i_2,l) = \mY_l(i_1,i_2),
% \end{equation}
% for all $(i_1,i_2) \in [p]^2$ and $l \in [L]$.

Assume there exist $r$ disjoint communities among $p$ nodes and $r_l$ disjoint communities among the $L$ layers. The multi-layer network community detection is modeled by the generalized asymmetric dTBM model~\eqref{eq:model_tensor}
\begin{equation}
    \bbE[\tY] = \tS \times_1 \mTheta \mM \times_2 \mTheta \mM \times_3 \mTheta_l \mM_l, 
\end{equation}
where $(\mtheta \in \bbR^p, \mM \in \{0,1\}^{p \times r})$ and $ (\mtheta_l \in \bbR^L, \mM_l \in \{0,1\}^{L \times r_l})$ are the degree heterogeneity and membership matrices corresponding to the community structure for $p$ nodes and $L$ layers, respectively. 
\vspace{0.2cm}

\paragraph{Gaussian higher-order clustering} Datasets in various fields such as medical image, genetics, and computer science are formulated as Gaussian tensors. One typical example is the multi-tissue gene expression dataset, which records the different gene expression in different individuals and different tissues. The dataset, denoted as $\tY \in \bbR^{p \times n \times t}$, consists of the expression data for $p$ genes of $n$ individuals in $t$ tissues.

Assume there exist $r_1, r_2, r_3$ disjoint clusters for $p$ genes, $n$ individuals, and $t$ tissues, respectively. We apply the generalized asymmetric dTBM model~\eqref{eq:model_tensor} 
\begin{equation}
    \bbE[\tY] = \tS \times_1 \mTheta_1 \mM_1 \times_2 \mTheta_2 \mM_2 \times_3 \mTheta_3 \mM_3, 
\end{equation}
where $\{(\mtheta_k, \mM_k)\}_{k=1}^3$ refers to the heterogeneity and membership for genes, individuals, and tissues.

}
% \begin{example}[Gaussian TBM]
% Let $\theta(i)= 1$ for all $i \in [p]$ and $\tE$ be a noise tensor with i.i.d.\ $N(0,\sigma^2)$ entries. Our dTBM reduces to a non-degree Gaussian TBM, which is widely used in previous clustering algorithms~\citep{wang2019multiway,chi2020provable,han2020exact}. The theoretical results in TBM serve as benchmarks for dTBM. 
% \end{example}

% \begin{example}[Binary dTBM]
% Consider a $K$-uniform hypergraph $H = (V,E)$, where $V = [p]$ collects the nodes with $r$ disjoint communities and $E$ collects all the $K$-way hyperedges. Let $\tY \in \{0,1\}^{p \times \cdots \times p}$ denote the adjacency tensor, where the entries encode the presence or absence of hyperdeges among $p$ nodes. Specifically, let $\tY(i_1,...,i_K) = 1$ if  $(i_1,...,i_K) \in E$, otherwise, $\tY(i_1,...,i_K) = 0$, for all $(i_1,\ldots,i_K) \in [p]^K$. The equation~\eqref{eq:model_tensor} models $\mathbb{E}\tY$ with unknown degree heterogeneity and subgaussianity parameter $\sigma^2 = 1/4$. 
% \end{example}

@@

\subsection{Identifiability under angle gap condition}
The goal of clustering is to estimate the partition function $z$ from model~\eqref{eq:model_tensor}. We use $\tP$ to denote the following parameter space for $(z,\tS,\mtheta)$,
\begin{align}\label{eq:family}
\tP=\bigg\{(z,\tS,\mtheta)\colon  \ \mtheta\in\mathbb{R}^p_{+},\ 
{c_1p\over r}\leq |z^{-1}(a)|\leq {c_2 p\over r}, c_3\leq \onormSize{}{\text{Mat}(\tS)_{a:}}\leq c_4, \onorm{\mtheta_{z^{-1}(a)}}_1=|z^{-1}(a)|, a\in[r]\bigg\},
\end{align}
\normalsize
where $c_i>0$'s are universal constants.  
%We briefly describe the rationale of the constraints in~\eqref{eq:family}. 
First, the entrywise positivity constraint on  $\mtheta\in\mathbb{R}^p_{+}$ is imposed to avoid sign ambiguity between entries in $\mtheta_{z^{-1}(a)}$ and $\tS$ {\color{blue} and thereof allow the trigonometric $\cos$ to describe the angle similarity in the following Assumption~\ref{assmp:min_gap} and Sub-algorithm~\hyperref[alg:main]{2} in Section~\ref{sec:alg}}. This constraint can be achieved without sacrificing model flexibility, by using a slightly larger dimension of $\tS$ in the factorization~\eqref{eq:model_tensor}; see Example~\ref{ex:positive}. Second, 
%the quantity $|z^{-1}(a)|$ denotes the number of nodes in $a$-th community. T
the constants $c_1, c_2$ in the $|z^{-1}(a)|$ bound assume the roughly balanced size across $r$ communities. Third, the constants $c_3, c_4 $ in the magnitude of $\text{Mat}(\tS)_{a:}$ requires no purely zero slide in $\tS$, so the core tensor $\tS$ is not trivially reduced to lower rank. Lastly, the $\ell_1$ normalization $\onormSize{}{\mtheta_{z^{-1}(a)}}_1=|z^{-1}(a)|$ is imposed to avoid the scalar ambiguity between $\mtheta_{z^{-1}(a)}$ and $\tS$. This constraint, again, incurs no restriction to model flexibility but makes our presentation cleaner.
